### 2015-10-05 paper/manuscript/plos_template.tex
proof paper

%There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
The potential for classical studies to yield insights into somatic mutation processes
were limited in several ways.
Due to limited sequencing throughput, most classical studies 
focused on a few cancer genes, such as TP53,
where high mutation frequencies could be expected. They then
contrasted mutation pattern profiles among different cancer types,
aggregating mutations across multiple individuals within the same cancer type to
yield sufficient mutations for analysis.
However, since many of the mutations in cancer genes are driver mutations causing cell proliferation,
the resultant mutation profiles are a biased representation of the underlying mutation process.
Furthermore, the paucity of mutation data made it effectively impossible to assess variation in mutation patterns among individuals.

@@

In this paper, we present a novel probabilistic approach 
to mutation signature modelling that addresses these limitations.
In brief, we first simplify the modelling of mutation signatures by decomposing them into separate ``mutation features".
For example, the substitution type is one feature; flanking bases are each another feature. We then exploit this 
decomposition by using a probabilistic model for signatures that assumes independence across features. This approach substantially reduces the number of parameters associated with
each signature, greatly facilitating the incorporation of additional relevant sequence context. 
For example, our approach can incorporate the two bases 3' and 5' of the substitution and transcription strand biases using only 18 parameters
per signature, compared with 3071 parameters per signature with current approaches. We demonstrate the benefits of this simplification in data analyses. These benefits include: more stable estimation of signatures from smaller samples, 
refinement of the detail and resolution of many mutation signatures, and possibly identification of novel signatures.

Assuming independence among features in a signature may initially seem unnatural. However, its use here is analogous to ``position weight matrix models" which have been highly
successful 
for modelling transcription factor binding motifs.
Indeed, an important second contribution of
our paper is to provide intuitive visual representations for mutation signatures,
analogous to the ``sequence logos" used for visualizing binding motifs.
Finally, we also highlight the close connection
between mutation signature models and the ``mixed-membership models'',
also known as``admixture models" \cite{pmid10835412} or ``latent Dirichlet allocation''
models \cite{Blei:2003} that are widely used in population genetics 
and document clustering applications.
These connections should be helpful for future elaboration of 
computational and statistical methods for cancer mutation signature detection.

@@

The first contribution of this paper is to suggest
a more parsimonious approach to modelling 
mutation signatures, with the benefit of producing both more stable estimates and more easily interpretable signatures.
In brief, we substantially reduce the number of parameters
per signature by breaking each mutation pattern into 
``features", and assuming independence
across mutation features. For example, consider
the case where a mutation pattern is defined by the substitution and its two flanking bases. We break this into three features (substitution,
3' base, 5' base), and characterize each mutation
signature by a probability distribution for each feature
(which, by our independence assumption, are multiplied
together to define a distribution on mutation patterns). Since the number of possible values for each feature is 6, 4, and 4 respectively this requires $5+3+3=11$ parameters instead of $96-1=95$ parameters. Furthermore, extending this model to account for $\pm n$ neighboring bases requires only $5 + 6n$ parameters instead of $6 \times 4^{2n} - 1$. For example,
 considering $\pm 2$ positions requires 17 parameters
 instead of 1535. Finally, incorporating
transcription strand as an additional feature adds just one parameter, instead of doubling the number of parameters.

@@

This generative model is summarized in Fig. \ref{methodOverview}.
This model is essentially a ``mixed-membership model", also known as an ``admixture model" \cite{pmid10835412} or ``latent Dirichlet allocation" \cite{Blei:2003}. For example,
the membership proportions for each sample are analogous to admixture proportions in an admixture model; the mutation signatures are analogous to 
populations, and the mutation signature-specific parameters are analogous to population-specific allele frequencies.

@@

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3071 parameters in the full model. After analyzing the data with various numbers of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (see S2 Text).

@@

\begin{figure}[h]
\caption{{\bf The mutation signatures for the UCUT data, and the results of down-sampling experiments.}
3072 elements in the full model mutation signatures were shown divided by 6 substitution patterns and strand directions.
(A, B) APOBEC and AA signature for the independent model.
(C, D) APOBEC and AA signature for the full model.
(E, F) APOBEC and AA signature stability (the mean cosine similarity for each down-sampling ratio).}
\label{UCUT}
\end{figure}

To investigate this further we performed down-sampling experiments.
Using the mutation signatures obtained using all 14717 substitutions as a gold standard, 
we assessed performance of the proposed method on down-sampled data consisting of $r\%$ of the original data, where  $r=(1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%)$.
To measure robustness we used the cosine similarity on the 
full dimensional vector space, which allows comparison between the full model and the independent model.
We repeated each down-sampling experiment 100 times for each model.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.
The selected values of $K$ are given in S2 Table.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius distances were $<$ $0.6$, where the Frobenius distance between two matrices ($\bm{F}_1$, $\bm{F}_2$) is $\sqrt{ \Tr  \bigl( (\bm{F}_1 -  \bm{F}_2 )  (\bm{F}_1 -  \bm{F}_2 )^t \bigr) }$,
where $\Tr$ means the trace of square matrices.
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm 2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. The 12 inferred signatures are highly consistent among cancer types except for B-cell lymphoma (see Fig. S3A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S3A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

@@

Although we have focused on point substitution mutations in this paper, 
many other types of mutations occur in cancer genomes, 
including insertions, deletions, double nucleotides substitutions, structural variations and copy number alterations \cite{meyerson2010advances, helleday2014mechanisms}.
Our framework could incorporate these additional mutation types, by 
summarizing them using appropriate mutation features.
In some cases, choice of appropriate features may need investigation. 
For example, longer deletions could be represented by the length of deletion and the adjacent bases; 
for short deletions (a few bases) it may be fruitful to include the actual deleted bases as part of the features.
In addition, it may be interesting to devise a probabilistic model of mutation signature somewhere between complete independence and non-independence assumption.
For example, assuming low-dimensional Markovian structure between adjacent bases as some transcription factor binding site modellings did \cite{pmid16108724}
may help improving representation power of mutation signatures keeping the number of parameters moderate.
%However, detailed investigation on what mutation features are practical is a future problem.

@@

Although we believe our new methods already provide useful gains compared with existing approaches,
the methods are perhaps even more important for their future potential to incorporate other contextual data, including epigenetic data, into mutation signature analysis. 
This is important, because local mutation densities are closely related to a number of genomic and epigenetic factors, 
such as GC content, repeat sequences, chromatin accessibility and modifications, and replication timing \cite{pmid22820252, pmid21953857, pmid23422670, pmid23770567}.
A recent study found that epigenetic information in the cell types of origin of the corresponding tumors is the most predictive \cite{pmid25693567} for local mutation densities. 
A growing range of epigenetic data from many cell types are now available, 
and it will be interesting to integrate these epigenetic factors into mutation signature analysis to help understand how these epigenetic factors influence DNA damage and repair mechanisms. 
Our work here provides a straightforward way to do this: epigenetic data can be simply added as features to the mutation signature.
In addition, a simulation study shows that the accuracy of mutation signature detection improves as we increase the number of mutation signatures (see Fig. S7).
We believe that the value and impact of our work, and specifically our proposed approach to modelling mutation signatures via independent features, will grow as more and more features
are incorporated into the analysis.

@@

The parameters $\{ \bm{f}_{k, l} \}$ and $\{ \bm{q}_i \}$
must be estimated from the available mutation data $\{ \bm{x}_{i,j} \}$. 
Here we adopt a simple approach that uses an EM-algorithm to maximise the likelihood.

>>>>

%There were several problems in these classical studies \cite{pmid12379884, pmid15748635}.
The potential for classical studies to yield insights into somatic mutation processes
was limited in several ways.
Due to limited sequencing throughput, most classical studies 
focused on a few cancer genes, such as TP53,
where high mutation frequencies could be expected. They then
contrasted mutation pattern profiles among different cancer types,
aggregating mutations across multiple individuals within the same cancer type to
yield sufficient mutations for analysis.
However, since many of the mutations in cancer genes are driver mutations causing cell proliferation,
the resultant mutation profiles are a biased representation of the underlying mutation process.
Furthermore, the paucity of mutation data made it effectively impossible to assess variation in mutation patterns among individuals.

@@

In this paper, we present a novel probabilistic approach 
to mutation signature modelling that addresses these limitations.
In brief, we first simplify the modelling of mutation signatures by decomposing them into separate ``mutation features".
For example, the substitution type is one feature; flanking bases are each another feature. We then exploit this 
decomposition by using a probabilistic model for signatures that assumes independence across features. This approach substantially reduces the number of parameters associated with
each signature, greatly facilitating the incorporation of additional relevant sequence context. 
For example, our approach can incorporate the two bases 3' and 5' of the substitution and transcription strand biases using only 18 parameters
per signature, compared with 3,071 parameters per signature with current approaches. We demonstrate the benefits of this simplification in data analyses. These benefits include: more stable estimation of signatures from smaller samples, 
refinement of the detail and resolution of many mutation signatures, and possibly identification of novel signatures.

Assuming independence among features in a signature may initially seem unnatural. However, its use here is analogous to ``position weight matrix models" which have been highly
successful 
for modelling transcription factor binding motifs.
Indeed, an important second contribution of
our paper is to provide intuitive visual representations for mutation signatures,
analogous to the ``sequence logos" used for visualizing binding motifs.
Finally, we also highlight the close connection
between mutation signature models and the ``mixed-membership models'',
also known as``admixture models" \cite{pmid10835412} or ``Latent Dirichlet Allocation''
models \cite{Blei:2003} that are widely used in population genetics 
and document clustering applications.
These connections should be helpful for future elaboration of 
computational and statistical methods for cancer mutation signature detection.

@@

The first contribution of this paper is to suggest
a more parsimonious approach to modelling 
mutation signatures, with the benefit of producing both more stable estimates and more easily interpretable signatures.
In brief, we substantially reduce the number of parameters
per signature by breaking each mutation pattern into 
``features", and assuming independence
across mutation features. For example, consider
the case where a mutation pattern is defined by the substitution and its two flanking bases. We break this into three features (substitution,
3' base, 5' base), and characterize each mutation
signature by a probability distribution for each feature
(which, by our independence assumption, are multiplied
together to define a distribution on mutation patterns). Since the number of possible values for each feature is 6, 4, and 4 respectively this requires $5+3+3=11$ parameters instead of $96-1=95$ parameters. Furthermore, extending this model to account for $\pm n$ neighboring bases requires only $5 + 6n$ parameters instead of $6 \times 4^{2n} - 1$. For example,
 considering $\pm 2$ positions requires 17 parameters
 instead of 1,535. Finally, incorporating
transcription strand as an additional feature adds just one parameter, instead of doubling the number of parameters.

@@

This generative model is summarized in Fig. \ref{methodOverview}.
This model is essentially a ``mixed-membership model", also known as an ``admixture model" \cite{pmid10835412} or ``Latent Dirichlet Allocation" \cite{Blei:2003}. For example,
the membership proportions for each sample are analogous to admixture proportions in an admixture model; the mutation signatures are analogous to 
populations, and the mutation signature-specific parameters are analogous to population-specific allele frequencies.

@@

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3,071 parameters in the full model. After analyzing the data with various numbers of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (see S2 Text).

@@

\begin{figure}[h]
\caption{{\bf The mutation signatures for the UCUT data, and the results of down-sampling experiments.}
3,072 elements in the full model mutation signatures were shown divided by 6 substitution patterns and strand directions.
(A, B) APOBEC and AA signature for the independent model.
(C, D) APOBEC and AA signature for the full model.
(E, F) APOBEC and AA signature stability (the mean cosine similarity for each down-sampling ratio).}
\label{UCUT}
\end{figure}

To investigate this further we performed down-sampling experiments.
Using the mutation signatures obtained using all 14,717 substitutions as a gold standard, 
we assessed performance of the proposed method on down-sampled data consisting of $r\%$ of the original data, where  $r=(1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%)$.
To measure robustness we used the cosine similarity on the 
full dimensional vector space, which allows comparison between the full model and the independent model.
We repeated each down-sampling experiment 100 times for each model.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.
The selected values of $K$ are given in S2 Table.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius distances were $<$ $0.6$, where the Frobenius distance between two matrices ($\bm{F}_1$, $\bm{F}_2$) is $\sqrt{ \Tr  \bigl( (\bm{F}_1 -  \bm{F}_2 )  (\bm{F}_1 -  \bm{F}_2 )^t \bigr) }$
($\Tr$ means the trace of square matrices).
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarize the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm 2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. 
The 12 inferred signatures were highly consistent among cancer types except for B-cell lymphoma (see Fig. S3A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S3A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

@@

Although we have focused on point substitution mutations in this paper, 
many other types of mutations occur in cancer genomes, 
including insertions, deletions, double nucleotides substitutions, structural variations and copy number alterations \cite{meyerson2010advances, helleday2014mechanisms}.
Our framework could incorporate these additional mutation types, by 
summarizing them using appropriate mutation features.
In some cases, choice of appropriate features may need investigation. 
For example, longer deletions could be represented by the length of deletion and the adjacent bases; 
for short deletions (a few bases) it may be fruitful to include the actual deleted bases as part of the features.
Also, it may be interesting to devise a probabilistic model for mutation signatures somewhere between complete independence and non-independence assumption, 
for example, using ideas analogous to those in \cite{pmid16108724} that uses a Markovian structure  for transcription  factor binding sites. 
This may help improve the modelling flexibility of mutation signatures while keeping the number of parameters moderate.

@@

Although we believe our new methods already provide useful gains compared with existing approaches,
the methods are perhaps even more important for their future potential to incorporate other contextual data, including epigenetic data, into mutation signature analysis. 
This is important, because local mutation densities are closely related to a number of genomic and epigenetic factors, 
such as GC content, repeat sequences, chromatin accessibility and modifications, and replication timing \cite{pmid22820252, pmid21953857, pmid23422670, pmid23770567}.
A recent study found that epigenetic information in the cell types of origin of the corresponding tumors is the most predictive \cite{pmid25693567} for local mutation densities. 
A growing range of epigenetic data from many cell types are now available, 
and it will be interesting to integrate these epigenetic factors into mutation signature analysis to help understand how these epigenetic factors influence DNA damage and repair mechanisms. 
Our work here provides a straightforward way to do this: epigenetic data can be simply added as features to the mutation signature. 
This has the potential to improve accuracy of signature detection (e.g. Fig S7), and to produce novel biological insights.
We believe that the value and impact of our work, and specifically our proposed approach to modelling mutation signatures via independent features, will grow as more and more features
are incorporated into the analysis.

@@

The parameters $\{ \bm{f}_{k, l} \}$ and $\{ \bm{q}_i \}$
must be estimated from the available mutation data $\{ \bm{x}_{i,j} \}$. 
Here we adopt a simple approach that uses an EM-algorithm to maximize the likelihood.
